---
title: "Example code for single-cell droplet quality control"
author: "Kevin Rue-Albrecht"
date: "05/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load packages
library(DropletUtils)
library(tidyverse)
library(cowplot)
library(Seurat)
library(scater)
library(scDblFinder)
library(DoubletFinder)
library(celda)
```

# Exercise

## Empty droplets

-   Import the raw matrix into R; use `DropletUtils`. Make a `SingleCellExperiment` object called `sce`.

```{r}
sce <- read10xCounts(
  samples = c(pbmc5k = "raw_feature_bc_matrix"), # Symlink to matrix folder within project directory
  col.names = TRUE
)
sce
```

-   Remove droplets without any count at all.

This will massively speed up downstream steps, while removing unquestionable empty droplets.

```{r}
zero.count <- colSums(counts(sce)) == 0 # Logical vector of barcodes with zero UMI counts
sce <- sce[, !zero.count] # Subset object to remove barcodes (columns) with zero UMI counts
sce
```

-   Run `DropletUtils::emptyDrops()`.

```{r}
set.seed(1)
out <- emptyDrops(
  m = sce,
  lower = 100,
  niters = 10000,
  assay.type = "counts"
)
out
```

**How do you read the output? Which droplets would you consider empty?**

> Answer: FDR \<0.01 (by convention)

-   Remove empty droplets from the data sets. Call the new object `sce_filtered_emptydrops`.

```{r}
non.empty.drops <- which(out$FDR < 0.01) # Index vector of non-empty drops
sce_filtered_emptydrops <- sce[, non.empty.drops] # Subset object to keep only non-empty drops (barcodes)
sce_filtered_emptydrops
```

## Bonus point

-   Draw plots comparing empty droplets and other droplets (e.g., library size, mitochondrial content, PCA).

```{r}
cell.status <- as.character(out$FDR < 0.01) %>% 
  replace(which(is.na(.)), "NA")

plot_data <- tibble(
 barcode = colnames(sce),
 library_size = colSums(counts(sce)),
 is_cell = factor(cell.status, levels = c("NA", "FALSE", "TRUE"))
)

ggplot(plot_data, aes(x = is_cell, y = library_size)) +
  geom_violin(fill = "orange") +
  scale_y_log10() +
  theme_cowplot()
```

```{r}
cell.status <- as.character(out$FDR < 0.01) %>% 
  replace(which(is.na(.)), "NA")

is.mt <- grep(rowData(sce)$Symbol, pattern = "^MT-")
percent.mt <- colSums(counts(sce)[is.mt, ]) / colSums(counts(sce)) * 100

plot_data <- tibble(
  barcode = colnames(sce),
  percent_mt = percent.mt,
  is_cell = factor(cell.status, levels = c("NA", "FALSE", "TRUE"))
)

ggplot(plot_data, aes(x = is_cell, y = percent_mt)) +
  geom_violin(fill = "orange") +
  theme_cowplot()
```

```{r}
# set.seed(1)
# sce <- logNormCounts(sce)
# sce <- runPCA(sce)
```

```{r}
# cell.status <- as.character(out$FDR < 0.01) %>% 
#   replace(which(is.na(.)), "NA")
# 
# plot_data <- tibble(
#   barcode = colnames(sce),
#   PC1 = reducedDim(sce, "PCA")$PC1,
#   PC2 = reducedDim(sce, "PCA")$PC2,
#   is_cell = factor(cell.status, levels = c("NA", "FALSE", "TRUE"))
# )
# 
# ggplot(plot_data) +
#   geom_point(aes(x = PC1, y = PC2, colour = is_cell)) +
#   theme_cowplot()
```

-   Import the filtered matrix for the same cells and compare their set of filtered cells to yours.

```{r}
sce_filtered_cellranger <- read10xCounts(
  samples = c(pbmc5k = "filtered_feature_bc_matrix"), # Symlink to matrix folder in project directory
  col.names = TRUE
)

all_barcodes <- colnames(sce)

compare_table <- tibble(
  emptydrops = all_barcodes %in% colnames(sce_filtered_emptydrops),
  cellranger = all_barcodes %in% colnames(sce_filtered_cellranger),
)
table(compare_table)
```

# Exercise

## Doublets

Continue using the data set after you excluded empty droplets.

-   Run `scDblFinder::scDblFinder()`.

```{r}
set.seed(1)
sce_scdblfinder <- scDblFinder(sce_filtered_emptydrops)
```

```{r}
colData(sce_scdblfinder)
```

```{r}
table(sce_scdblfinder$scDblFinder.class)
```

**How do you read the output? Which droplets would you consider doublets?**

> Answer: Droplets with `scDblFinder.class` of doublet

-   Remove doublets from the data set. Call the new object `sce_filtered_scdblfinder`.

```{r}
is.singlet <- sce_scdblfinder$scDblFinder.class == "singlet"
sce_filtered_scdblfinder <- sce_scdblfinder[, is.singlet]
sce_filtered_scdblfinder
```

## Bonus point

-   Draw plots comparing doublets and other droplets. (e.g., library size, mitochondrial content, PCA)

```{r}
plot_data <- tibble(
  barcode = colnames(sce_scdblfinder),
  library_size = colSums(counts(sce_scdblfinder)),
  droplet_type = sce_scdblfinder$scDblFinder.class
)

ggplot(plot_data) +
  geom_violin(aes(x = droplet_type, y = library_size), fill = "orange") +
  theme_cowplot()
```

```{r}
is.mt <- grep(rowData(sce_scdblfinder)$Symbol, pattern = "^MT-")
percent.mt <- colSums(counts(sce_scdblfinder)[is.mt, ]) / colSums(counts(sce_scdblfinder)) * 100

plot_data <- tibble(
  barcode = colnames(sce_scdblfinder),
  percent_mt = percent.mt,
  droplet_type = sce_scdblfinder$scDblFinder.class
)

ggplot(plot_data) +
  geom_violin(aes(x = droplet_type, y = percent_mt), fill = "orange") +
  theme_cowplot()
```

```{r}
set.seed(1)
sce_scdblfinder <- logNormCounts(sce_scdblfinder) 
sce_scdblfinder <- runPCA(sce_scdblfinder)

plot_data <- tibble(
  barcode = colnames(sce_scdblfinder),
  PC1 = reducedDim(sce_scdblfinder, "PCA")[, "PC1"],
  PC2 = reducedDim(sce_scdblfinder, "PCA")[, "PC2"],
  droplet_type = sce_scdblfinder$scDblFinder.class
)

ggplot() +
  geom_point(data = plot_data %>% filter(droplet_type == "singlet"),
             mapping = aes(x = PC1, y = PC2),
             size = 0.5,
             colour = "black") +
  geom_point(data = plot_data %>% filter(droplet_type == "doublet"),
             mapping = aes(x = PC1, y = PC2),
             size = 0.5,
             colour = "red") +
  theme_cowplot()
```

```{r}
set.seed(1)
sce_scdblfinder <- runUMAP(sce_scdblfinder, dimred = "PCA", n_dimred = 20)

plot_data <- tibble(
  barcode = colnames(sce_scdblfinder),
  UMAP1 = reducedDim(sce_scdblfinder, "UMAP")[, 1],
  UMAP2 = reducedDim(sce_scdblfinder, "UMAP")[, 2],
  droplet_type = sce_scdblfinder$scDblFinder.class
)

ggplot() +
  geom_point(data = plot_data %>% filter(droplet_type == "singlet"),
             mapping = aes(x = UMAP1, y = UMAP2),
             size = 0.5,
             colour = "black") +
  geom_point(data = plot_data %>% filter(droplet_type == "doublet"),
             mapping = aes(x = UMAP1, y = UMAP2),
             size = 0.5,
             colour = "red") +
  theme_cowplot()
```

-   Compare results with the output of `chris-mcginnis-ucsf/DoubletFinder`

    -   You will need to make a `Seurat` object.

```{r, message = FALSE}
## Pre-process Seurat object (standard)
seu <- CreateSeuratObject(
  counts = counts(sce_filtered_emptydrops),
  project = "pbmc5k",
  assay = "RNA"
)

# is.mt <- grep(rowData(sce_filtered_emptydrops)$Symbol, pattern = "^MT-")
# percent.mt <- colSums(counts(sce_filtered_emptydrops)[is.mt, ]) / colSums(counts(sce_filtered_emptydrops)) * 100
# seu$percent_mt <- percent.mt
# seu <- subset(seu, subset = nCount_RNA > 2500 & nFeature_RNA > 1500 & percent_mt < 15)

seu <- NormalizeData(seu)
seu <- ScaleData(seu)

seu <- FindVariableFeatures(seu,
                            selection.method = "vst",
                            nfeatures = 2000
                            )

seu <- RunPCA(seu,
              features = VariableFeatures(seu)
              )

seu <- RunUMAP(seu,
               reduction = "pca",
               dims = 1:20
               )

seu <- FindNeighbors(seu,
                     k.param = 20,
                     reduction = "pca",
                     dims = 1:20,
                     assay = "RNA"
                     )

seu <- FindClusters(seu,
                    resolution = 0.5,
                    algorithm = 1
                    )
```

```{r, message=FALSE}
## pK Identification (no ground-truth) 
sweep.res.list <- paramSweep_v3(seu, PCs = 1:20, sct = FALSE)
sweep.stats <- summarizeSweep(sweep.res.list, GT = FALSE)
bcmvn <- find.pK(sweep.stats)
```

```{r}
bcmvn
```

```{r}
# ggplot() +
#   
#   
#   
```

```{r}
## Homotypic Doublet Proportion Estimate 
homotypic.prop <- modelHomotypic(Idents(seu)) 
nExp_poi <- round(0.05 * nrow(seu@meta.data))  ## Assuming 5% doublet formation rate - tailor for your dataset
nExp_poi.adj <- round(nExp_poi * (1 - homotypic.prop))
```

```{r, message=FALSE}
## Run DoubletFinder with varying classification stringencies 
seu <- doubletFinder_v3(seu, PCs = 1:20, pN = 0.25, pK = 0.09, nExp = nExp_poi, reuse.pANN = FALSE, sct = FALSE)
```

```{r}
table(seu@meta.data$DF.classifications_0.25_0.09_264)
```

```{r, message=FALSE}
seu <- doubletFinder_v3(seu, PCs = 1:10, pN = 0.25, pK = 0.09, nExp = nExp_poi.adj, reuse.pANN = "pANN_0.25_0.09_264", sct = FALSE)
```

```{r}
table(seu@meta.data$DF.classifications_0.25_0.09_229)
```

```{r}
doublets_table <- data.frame(
  scdblfinder = sce_scdblfinder$scDblFinder.class,
  doubletfinder = seu@meta.data$DF.classifications_0.25_0.09_264
)
table(doublets_table)
```

# Exercise

## Ambient RNA

-   Run `celda::decontX()` (continue using the data after excluding empty droplets and doublets).

```{r, message=FALSE}
set.seed(1)
sce_decontX <- decontX(sce_filtered_scdblfinder)
```

```{r}
sce_decontX
```

```{r}
head(colData(sce_decontX))
```

-   Visualize the UMAP layout created by `decontX`; use `plotDimReduceCluster()`.

```{r}
umap <- plotDimReduceCluster(
  x = colData(sce_decontX)$decontX_clusters,
  dim1 = reducedDim(sce_decontX, "decontX_UMAP")[, 1],
  dim2 = reducedDim(sce_decontX, "decontX_UMAP")[, 2]
)
umap
```

-   Visualize the percentage of contamination estimated by `decontX` in each cell. Use the `plotDecontXContamination()` function.

```{r}
plotDecontXContamination(sce_decontX)
```

**Does any cluster display higher (estimated) levels of ambient RNA?**

> Answer: Yes

-   Visualize the log-normalised expression of marker genes on a UMAP layout faceted by gene; use `plotDimReduceFeature()`.

**Note:** For instance, use `CD3D`, `CD3E`, `GNLY`, `LYZ`, `S100A8`, `S100A9`, `CD79A`, `CD79B`, and `MS4A1`.

```{=html}
<!--
Source: https://bioconductor.org/packages/devel/bioc/vignettes/celda/inst/doc/decontX.pdf
(Section 5.3)
We will use:
- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes
-->
```
```{r}
rownames(sce_decontX) <- uniquifyFeatureNames(rowData(sce_decontX)$ID, rowData(sce_decontX)$Symbol)

sce_decontX <- logNormCounts(sce_decontX, assay.type = "counts", name = "logcounts")
sce_decontX <- logNormCounts(sce_decontX, assay.type = "decontXcounts", name = "logdecontXcounts")

plotDimReduceFeature(
  x = assay(sce_decontX, "logdecontXcounts"),
  features = c("CD3D", "CD3E", "GNLY", "LYZ", "S100A8", "S100A9", "CD79A", "CD79B", "MS4A1"),
  dim1 = reducedDim(sce_decontX, "decontX_UMAP")[, 1],
  dim2 = reducedDim(sce_decontX, "decontX_UMAP")[, 2]
)
```

-   Visualise the percentage of cells with detectable expression of marker genes before and after decontamination as a bar plot; use `plotDecontXMarkerPercentage()`.

```{r}
markers <- list(
  T_cell = c("CD3D", "CD3E"),
  B_cell = c("CD79A", "CD79B", "MS4A1"),
  NK_cell = c("GNLY"),
  mono = c("LYZ", "S100A8", "S100A9"),
  mega = c("PPBP")
)
cellTypeMappings <- list(
  T_cell = 1,
  B_cell = 3,
  NK_cell = 5,
  mono = 2,
  mega = c(6, 7, 8)
)

plotDecontXMarkerPercentage(
  sce_decontX,
  markers = markers,
  groupClusters = cellTypeMappings,
  assayName = c("counts", "decontXcounts")
)
```

-   Visualize the log-normalised expression of marker genes before and after decontamination as a violin plot; use `plotDecontXMarkerExpression()`.

```{r fig.width = 10, fig.height = 10}
plotDecontXMarkerExpression(
  sce_decontX,
  markers = markers,
  groupClusters = cellTypeMappings,
  assayName = c("logcounts", "logdecontXcounts"),
)
```
